---
title: "Stock Predictor — Model Report"
subtitle: "Next-day direction prediction from technical indicators (XGBoost)"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9)

suppressPackageStartupMessages({
  library(ggplot2)
  library(jsonlite)
  library(dplyr)
  library(knitr)
})

features_path <- file.path("..", "data", "features", "features.csv")
eval_log_path <- file.path("..", "models", "eval_log.json")

if (!file.exists(features_path) || !file.exists(eval_log_path)) {
  stop("Pipeline outputs not found. Run `python run_pipeline.py` from the project root before rendering this report.")
}

features <- read.csv(features_path, stringsAsFactors = FALSE)
features$Date <- as.Date(features$Date)
features <- features[order(features$Date), ]

eval_log <- fromJSON(eval_log_path)
```

## Approach

This report summarizes a model that predicts whether a stock's closing
price will be **higher tomorrow than it closed today** — a binary
up/down classification, not a price forecast.

**Pipeline:**

1. **Data collection (Python / yfinance):** `r nrow(features)` daily bars (after feature engineering) are downloaded for the ticker (5 years by default).
2. **Feature engineering (R / TTR):** SMA(20, 50), EMA(20), RSI(14), MACD(12, 26, 9) with signal & histogram, Bollinger Bands(20, 2σ), and ATR(14) are computed from the raw OHLCV series. The target label is `1` if the *next* trading day's close is higher than today's close, else `0`. Rows with indicator warm-up `NA`s are dropped.
3. **Model training (Python / XGBoost):** a gradient-boosted tree classifier is trained on a **chronological** 80/20 split (train on the earlier 80% of history, test on the most recent 20%) so no future information leaks into training — a random shuffle would be inappropriate for time-series data.
4. **Serving:** the trained model is exposed through a Shiny dashboard for live predictions and this R Markdown report for offline evaluation.

## Data Summary

```{r data-summary}
data.frame(
  Metric = c("Total rows (post feature engineering)", "Date range", "Up days", "Down days", "Up-day %"),
  Value = c(
    nrow(features),
    paste(min(features$Date), "to", max(features$Date)),
    sum(features$Target == 1),
    sum(features$Target == 0),
    sprintf("%.1f%%", 100 * mean(features$Target))
  )
) %>% kable(col.names = c("Metric", "Value"))
```

## Model Performance

```{r perf-summary}
data.frame(
  Metric = c("Training rows", "Test rows", "Train date range", "Test date range",
             "Test accuracy", "Majority-class baseline accuracy", "Uplift over baseline"),
  Value = c(
    eval_log$n_train,
    eval_log$n_test,
    paste(eval_log$train_date_range, collapse = " to "),
    paste(eval_log$test_date_range, collapse = " to "),
    sprintf("%.2f%%", eval_log$accuracy * 100),
    sprintf("%.2f%%", eval_log$baseline_accuracy * 100),
    sprintf("%+.2f pp", 100 * (eval_log$accuracy - eval_log$baseline_accuracy))
  )
) %>% kable(col.names = c("Metric", "Value"))
```

The **majority-class baseline** always predicts the most common label seen
in training (usually "Up", since equities drift upward more often than
not). A useful model should beat this baseline — if it doesn't, the
technical indicators aren't adding predictive signal beyond the market's
natural upward drift.

### Confusion Matrix

```{r confusion-matrix}
cm <- eval_log$confusion_matrix$matrix
rownames(cm) <- paste("Actual", eval_log$confusion_matrix$labels)
colnames(cm) <- paste("Predicted", eval_log$confusion_matrix$labels)
kable(cm)
```

### Classification Report

```{r classification-report}
rep <- eval_log$classification_report
rows <- c("Down", "Up")
report_df <- do.call(rbind, lapply(rows, function(r) {
  data.frame(
    Class = r,
    Precision = sprintf("%.3f", rep[[r]]$precision),
    Recall = sprintf("%.3f", rep[[r]]$recall),
    F1 = sprintf("%.3f", rep[[r]][["f1-score"]]),
    Support = rep[[r]]$support
  )
}))
kable(report_df)
```

### Feature Importance

```{r feature-importance, fig.height=5}
fi <- eval_log$feature_importance %>%
  arrange(desc(importance)) %>%
  head(12)
fi$feature <- factor(fi$feature, levels = rev(fi$feature))

ggplot(fi, aes(x = feature, y = importance)) +
  geom_col(fill = "#2563eb") +
  coord_flip() +
  labs(x = NULL, y = "Importance (gain)", title = "Top 12 Features by Importance") +
  theme_minimal(base_size = 13)
```

## Price & Indicator Plots

```{r price-plot, fig.height=4.5}
ggplot(features, aes(x = Date)) +
  geom_line(aes(y = Close, color = "Close")) +
  geom_line(aes(y = SMA_20, color = "SMA 20")) +
  geom_line(aes(y = SMA_50, color = "SMA 50")) +
  scale_color_manual(values = c("Close" = "#1f2937", "SMA 20" = "#2563eb", "SMA 50" = "#f59e0b")) +
  labs(title = "Closing Price with Moving Averages", x = NULL, y = "Price ($)", color = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

```{r rsi-plot, fig.height=3}
ggplot(features, aes(x = Date, y = RSI_14)) +
  geom_line(color = "#7c3aed") +
  geom_hline(yintercept = c(30, 70), linetype = "dashed", color = "#9ca3af") +
  labs(title = "RSI (14)", x = NULL, y = "RSI") +
  theme_minimal(base_size = 13)
```

```{r macd-plot, fig.height=3.5}
ggplot(features, aes(x = Date)) +
  geom_col(aes(y = MACD_hist), fill = "#a5b4fc") +
  geom_line(aes(y = MACD, color = "MACD")) +
  geom_line(aes(y = MACD_signal, color = "Signal")) +
  scale_color_manual(values = c("MACD" = "#2563eb", "Signal" = "#dc2626")) +
  labs(title = "MACD (12, 26, 9)", x = NULL, y = NULL, color = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

## Limitations & Disclaimer

- This model predicts **direction only**, not magnitude — a 0.1% up day
  and a 5% up day are labeled identically.
- Technical indicators are lagging by construction; predictive power over
  a pure drift baseline is often small and can vary a lot by ticker and
  regime.
- Markets are noisy and non-stationary: a model that performed well on
  historical test data is not guaranteed to perform well going forward.
- **This project is for educational purposes only and does not constitute
  financial advice.** Do not use it to make real trading decisions.

---

*Generated automatically from `data/features/features.csv` and
`models/eval_log.json` — re-run `python run_pipeline.py` and re-knit this
report to refresh with new data.*
